---
title: "R_Coursework_2022"
author: "Leo Jones"
date: "2022-12-08"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table,glue)
library(heplots)
library(stringi)
library(stringr)
setwd("C:/Users/Leo Jones/Downloads")
```

Q1. Reading the data into R:
```{r q1}
dat <- read.table("prob.txt")
```


Q2. Formatting table:
```{r q2}
dat <- paste(dat,collapse =' ')
dat <- gsub("[[:punct:]]", "", gsub("-"," ",dat)) #removing punctuation
dat<- data.frame(strsplit(dat, "[[:space:]]")) #splitting into words
colnames(dat)<- ("Words")
head(dat)
```


Q3. Creating a letter count:
```{r q3}
dat$Count <- nchar(dat$Words)
hist(dat$Count, xlab='Word Length', main='Frequency of lengths of words')
```

Q4. New column for word length:
```{r q4}
dat$State <- ifelse(dat$Count<4, '0', '1')
```

Q5. Summing over this column:
```{r q5}
paste(sum(strtoi(dat$State)), ' long words out of ', nrow(dat), '\n', nrow(dat)- sum(strtoi(dat$State)), ' short words out of ', nrow(dat))
```

Q6. Counting pairs:
```{r q6}
dat$Pair <- mapply(dat$State, shift(dat$State, 1, 2, 'lead'), FUN = function(a,b){paste(a,b)} ) #this adds a column shifted by one value and then compares the two columns
n11 <- nrow(subset(dat, dat$Pair=='1 1')) 
n10 <- nrow(subset(dat, dat$Pair=='1 0')) 
n01 <- nrow(subset(dat, dat$Pair=='0 1')) 
n00 <- nrow(subset(dat, dat$Pair=='0 0')) 
```

Q7. Estimating proportions based on the values obtained in (5):
```{r q7}
paste('The expected proportion of 11 is', signif((122/193)^2,4),'\n The expected proportion of 10 is ', signif(122*71/193^2, 4),'\n The expected proportion of 01 is ', signif(122*71/193^2,4),'\n The expected proportion of 00 is ', signif((71/193)^2,4))

```

Q8. Log likelihood statistic: \n
To avoid overflow errors, I calculated the likelihoods already under logarithms. Due to this we can ignore any constant term (from likelihood being directly proportional to the sum of p/q's) as it will cancel out when you subtract the q statistic from the p statistic. I've defined the log likelihood for p and q as functions as this will be useful later.
```{r q8}
loglq <- function(a,b,c,d){
  return( ((2*a)+b+c)*log(122) + ((2*d)+b+c)*log(71) - 384*log(193))
}
#likeq <- 132*log(122) + 111*log(122*71) + 30*log(71) - 384*log(193)
loglp <- function(a,b,c,d){
  return(a*log(a) + b*log(b) + c*log(c) + d*log(d) - 192*log(192))
}
likepq <- loglp(n11,n10,n01,n00) - loglq(n11,n10,n01,n00)

```

Q9. Log likelihood for random permutations: \n
I created a function that generates random permutations and returns the log likelihood statistic. I then displayed this using a density plot.
```{r q9}
randperm <- function(){
  perm <- sample(dat$State, size=193, replace=FALSE) #smaple function randomizes permutation
  pairs <- paste(perm , shift(perm, 1, 2, 'lead')) #shifting and concatenating
  a=sum(pairs=='1 1')
  b=sum(pairs=='1 0')
  c=sum(pairs=='0 1')
  d=sum(pairs=='0 0')
  return(loglp(a,b,c,d)-loglq(a,b,c,d)) #returning log likelihood statistic 
  
}

s=replicate(10000,(randperm()))
plot(density(c(s)), main='Density of Log Likelihoods', xlab= 'Log likelihood statistic') #replicate runs this function 10000 times
paste('Mean: ',round(mean(s),4),'\n Standard deviation: ',round(sd(s),4))
```
\n
We get graph highly concentrated around one value, with a mean of ~0.5 and sd ~ 0.7. Most of our values lie  between 0 and 1 with all of them positive.


Q10. QQ-plot with twice our log likelihoods to a chi- squared distribution \n We have one degree of freedom as we have two free parameters estimated under the alternative hypothesis and three under the null hypothesis.
```{r q10}
qqPlot(c(s)*2, distribution ='chisq', df=1, ylab='Sample Quantities', xlab='Theoretical Quantities', main='Chi-squared QQ plot')
#plotting qqplot with chi-squared 1 degree of freedom

```

Q11. Comparing this to our first log likelihood:
```{r q11}
paste(quantile(c(replicate(10000,(randperm())))*2, probs=(0.95))) #95th percentile of the doubled log likelihood

paste(qchisq(0.95,1)) #95th percentile of the chi-squared distribution for comparison
```
We need to consider half of this value ~1.9 as we doubled our log likelihood. At 5% significance, 1.9<6.0 (from Q8) and we therefore reject the null hypothesis.

Q12. Creating the matrix of transition probabilities using proportions in (6):
```{r q12}
tmatrix <- matrix(c(15/71,56/71,55/121,66/121), nrow=2) #these are eaily figured out from a diagram
print(tmatrix)
```

Q13. Function that simulates a draw of length m:
```{r q13}
sim <- function(m,x_0){
  c <- list()
  if(m>1){
    x_1=ifelse(x_0==0, ifelse(sample(1:71,1)>15,1,0), ifelse(sample(1:121,1)>55,0,1))
    #using nested ifs to calculate x_n+1 from x_n
    c=sim(m-1,x_1)} #generating the draw recursively 
  return(append(c,x_0,after=0))
}

paste(sim(10,0)) #a random draw of length 10, starting at 0
```

Q14. Simulating n=100 realizations and using these to estimate the transition probabilities:
```{r q14}
real <- function(m,x_0){
  return(lapply(str_count(stri_join_list(sim(m,x_0),collapse = ''),paste0("(?=",c("00","01","10","11"),")")), FUN= function(a){round(a/(m-1),4)}))
#this function runs the simulation once and returns the estimated transition probabilities, to 4sf
}

estimates <- data.frame(t(replicate(100,real(100,0)))) #storing the results of 100 estimations into a data frame 
head(estimates)

X1=sum(as.numeric(estimates$X1))/100
X2=sum(as.numeric(estimates$X2))/100
X3=sum(as.numeric(estimates$X3))/100
X4=sum(as.numeric(estimates$X4))/100 #I know I can do this more efficiently using apply but it wasn't working :(
paste(X1,',',X2,',',X3,',',X4/100) #means of the columns

#the pairwise correlations are as follows
paste(cor(as.numeric(estimates$X1),as.numeric(estimates$X2)))
paste(cor(as.numeric(estimates$X1),as.numeric(estimates$X3)))
paste(cor(as.numeric(estimates$X1),as.numeric(estimates$X4)))
paste(cor(as.numeric(estimates$X2),as.numeric(estimates$X3)))
paste(cor(as.numeric(estimates$X4),as.numeric(estimates$X2)))
paste(cor(as.numeric(estimates$X3),as.numeric(estimates$X4)))

```
The means for my random realizations are similar to that of the transition matrix, implying that the estimators are unbiased.


Q15.
```{r q15}

```
